Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-04-30 doc/paper.Rmd
line breaks at end of sentence
hyphens, n-dashes, m-dashes

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015]. Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat. Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others? -->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others? -->. However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others? -->. This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates. For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.

@@

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome. For instance, <!--#parts of th      e Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013]. These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions. In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].

Add here what is known about frag and drought effects - without setting it up so people expect a pval for an intearaciton tern

@@

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).

@@

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004]. Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
This study uses data from 1-ha fragments and Continuous Forest sites.

@@

<!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).  -->

@@

<!-- re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->

@@

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodChapterSmoothers] and Gasparrini et al. [-@gasparriniPenalizedFrameworkDistributed2017] for more detail). Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->. Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size.
A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function .

In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013]. <!--# more here -->. Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years). Using a decadal dataset of plant demography, we show that some - but not all - results from short-term studies can be extrapolated to longer times scales. Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

@@

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002]. In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002]. While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher. The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences. However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF). These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002]. However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest. Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

@@

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including tropical understory herbs [@wrightSeasonalDroughtSoil1992]. For instance, @westerbandEarlyLifeConditions2017: found that *Heliconia* grew more at low than high precipitation, while @stiles1975: observed reductions in flowering as a result of drought. Despite this, models of population dynamics have rarely included the effects of environmental drivers. This has largely been due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons. To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.

While it appears that delayed effects of climate on demographic vital rates are ubiquitous, however, the extent to which they vary spatially or with habitat remains an open question [@eversLaggedDormantSeason2021]. Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates. For example, [delete this phrase? the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically relevant.
For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences. Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994]. Interestingly, we observed the opposite effect - drought two dry seasons prior to a census actually increased the observed probability of flowering. While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, and there do not appear to be subsequent decreases in growth or survival following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering is associated with increased reproduction and growth in the following year **(is this an acceptable colloquial way to interpret the covariate effect?)**.

Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
For example, the mortality rate of tropical trees can spike in the year following a severe drought.
The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments (**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?**)**.
Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon). Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.

@@

Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
A major consequence of forest fragmentation is reported alteration in microclimate near forest edges [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013]. It is possible that increases in temperature, evaporative drying, wind speed, and solar radiation commonly found near forest edges exacerbated the effects of extremes in SPEI on *H. acuminata* growth and survival in fragments compared to continuous forest. It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments. Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.

>>>>

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others? -->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.

@@

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
For instance, <!--# parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50--100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable---particularly for plants, which are unable to move in response to changes in local conditions.
In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].

<!--# Add here what is known about frag and drought effects - without setting it up so people expect a pval for an interaction term -->

@@

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900--2300 mm.
There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).

@@

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest (n=6), 10 ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
This study uses data from 1-ha fragments and Continuous Forest sites.

@@

<!--# I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).  -->

@@

<!--# re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->

@@

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see @woodChapterSmoothers and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->.
Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size.
A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.

In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in @marraPracticalVariableSelection2011.

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
<!--# more here -->.
Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
Using a decadal dataset of plant demography, we show that some---but not all---results from short-term studies can be extrapolated to longer times scales.
Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

@@

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF).
These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest.
Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

@@

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including tropical understory herbs [@wrightSeasonalDroughtSoil1992].
For instance, @westerbandEarlyLifeConditions2017: found that *Heliconia* grew more at low than high precipitation, while @stiles1975: observed reductions in flowering as a result of drought.
Despite this, models of population dynamics have rarely included the effects of environmental drivers.
This has largely been due to the challenge---both ecologically and statistically---of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.

While it appears that delayed effects of climate on demographic vital rates are ubiquitous, however, the extent to which they vary spatially or with habitat remains an open question [@eversLaggedDormantSeason2021].
Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
For example, [delete this phrase?
the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically relevant.
For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 <!--# weeks? --> prior to the appearance of infloresences.
Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
Interestingly, we observed the opposite effect - drought two dry seasons prior to a census actually increased the observed probability of flowering.
While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, and there do not appear to be subsequent decreases in growth or survival following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering is associated with increased reproduction and growth in the following year **(is this an acceptable colloquial way to interpret the covariate effect?)**.

Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
For example, the mortality rate of tropical trees can spike in the year following a severe drought.
The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments <!--#  (question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?)-->.
Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.

@@

Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
A major consequence of forest fragmentation is reported alteration in microclimate near forest edges [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
It is possible that increases in temperature, evaporative drying, wind speed, and solar radiation commonly found near forest edges exacerbated the effects of extremes in SPEI on *H. acuminata* growth and survival in fragments compared to continuous forest.
It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1-ha fragments.
Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.
